Solution of hydrodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2014-06-20 pituba/article/article_spe.tex
continua

Reservoir fluids are composed of oil and gas pseudo-components and water component. There can be three fluid phases: oil, gas and (liquid) water. Oil phase may be composed of oil and gas components, gas phase can be composed uniquely by gas component and water phase uniquely by water component. The only allowed mass transfer is of gas component between oil and gas phases, i.e., composition of gas phase is always that of a dry gas.

@@

Mass balance equations for each component are:

\begin{itemize}
\item water component:
\begin{equation}\label{eq: Sw1}
\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV + \oint_{\partial \Omega} \rho_w \left( \vec{u}_w \cdot \vec{dS} \right) = \rho_w^{ST}Q_w^{ST} \, ;
\end{equation}

\item oil component:
\begin{equation}\label{eq: So1}
\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV + \oint_{\partial \Omega} \rho_o Y_o \left(\vec{u}_o \cdot \vec{dS} \right)= \rho_o^{ST}Q_o^{ST} \, ;
\end{equation}

\item gas component:
\begin{equation}\label{eq: Sg1}
\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV + \oint_{\partial \Omega} \left( \rho_o Y_g \vec{u}_o + \rho_g \vec{u}_g\right) \cdot \vec{dS}= \rho_g^{ST}Q_g^{ST} \, .
\end{equation}
\end{itemize}
Mass concentrations of oil and gas components in oil phases are denoted as $Y_o$ and $Y_g$, respectively.

Phase saturation is defined as its volumetric fraction of the pore space:
\begin{equation}
S_\alpha = \frac{\phi_\alpha}{\phi} \, ,
\end{equation}
where $\phi_\alpha$ is the volumetric fraction of a fluid phase $\alpha$ and $\phi$ is the porosity. Clearly, all saturations sum up to unity. 
\begin{equation}
S_o + S_g +S_w = 1 \, .
\end{equation}

@@

Derivation of the material balance equation consists in assuming no fluid motion and in integrating black-oil equations in reservoir volume. As it will be shown, the integration process requires all quantities to be either spatially averaged or assumed constant.

@@

Black-oil equations become:
\begin{itemize}
\item water component:
\begin{equation}\label{eq: Sw2}
\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV = \rho_w^{ST}Q_w^{ST} \, ;
\end{equation}

\item oil component:
\begin{equation}\label{eq: So2}
\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV = \rho_o^{ST}Q_o^{ST} \, ;
\end{equation}

\item gas component:
\begin{equation}\label{eq: Sg2}
\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV = \rho_g^{ST}Q_g^{ST} \, .
\end{equation}
\end{itemize}
%
\subsubsection{Averaging quantities}

@@

All quantities are initially functions of space and time and they become functions only of time after being averaged in space:

@@

\begin{itemize}
\item water component:
\begin{equation}\label{eq: Sw3}
\frac{\partial}{\partial t} \left( \bar{\rho_w}\bar{S}_w\bar{\phi} \Omega \right) = \rho_w^{ST}Q_w^{ST}
\end{equation}

@@

\item gas component:
\begin{equation}\label{eq: Sg3}
\frac{\partial}{\partial t} \left[\left( \bar{\rho}_o\bar{Y}_g \bar{S}_o + \bar{\rho}_g \bar{S}_g \right) \bar{\phi} \Omega\right] = \rho_g^{ST}Q_g^{ST} \, .
\end{equation}
\end{itemize}

@@

\begin{equation}\label{eq: Sw4}
\frac{\partial}{\partial t} \left(\frac{\bar{S}_w V_p}{B_w} \right) = Q_w^{ST} \, ,
\end{equation}

@@

\begin{equation}\label{eq: Sg4}
\frac{\partial}{\partial t} \left[ V_p \left(\frac{\bar{S}_g}{B_g} + \frac{\bar{S}_o R_s}{B_o} \right) \right] = Q_g^{ST} \, ,
\end{equation}

>>>>

%Phase saturation is defined as the fraction of the occupied pore volume:
%\begin{equation}
%S_\alpha = \frac{\phi_\alpha}{\phi} \, ,
%\end{equation}
%where $\phi_\alpha$ is the volumetric fraction of a fluid phase $\alpha$ and $\phi$ is the porosity. Clearly, all saturations sum up to unity. 
%\begin{equation}
%S_o + S_g +S_w = 1 \, .
%\end{equation}

We assume that the reservoir is a connected volume where hydrocarbon saturation is nonzero. Reservoir fluids are composed of oil and gas pseudo-components and water component. There can be three fluid phases: oil, gas and (liquid) water. Oil phase may be composed of oil and gas components, gas phase is composed uniquely by gas component and water phase uniquely by water component. The only allowed mass transfer is of gas component between oil and gas phases, i.e., gas phase composition is always that of a dry gas.

@@

Mass balance equations for each component read:

\begin{itemize}
\item water component:
\begin{equation}\label{eq: Sw1}
\frac{\partial}{\partial t} \left( \rho_w S_w \phi \right) +\nabla \cdot \left( \rho_w \vec{u}_w \right) = 0 ;
\end{equation}

\item oil component:
\begin{equation}\label{eq: So1}
\frac{\partial}{\partial t}\left( \rho_o Y_o S_o \phi \right) + \nabla \cdot \left( \rho_o Y_o \vec{u}_o \right) = 0 \, ;
\end{equation}

\item gas component:
\begin{equation}\label{eq: Sg1}
\begin{split}
\frac{\partial}{\partial t} \left( \rho_o Y_g S_o + \rho_g S_g\right) + \nabla  \cdot \left( \rho_o Y_g \vec{u}_o + \rho_g \vec{u}_g\right) = 0\, .
\end{split}
\end{equation}
\end{itemize}
Mass concentrations of oil and gas components in oil phase are denoted by $Y_o$ and $Y_g$, respectively.

Initial conditions are of no fluid motion and boundary conditions are of no normal flow at all boundaries except for interfaces with wells or aquifers.

@@

Derivation of the material balance equation consists in integrating black-oil equations in reservoir volume.

@@

Integrating Eqs.~\eqref{eq: Sw1}, \eqref{eq: So1} and \eqref{eq: Sg1} in volume and applying divergence theoreom gives:
\begin{itemize}
\item water component:
\begin{equation}\label{eq: Sw2}
\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV = - \oint_{\partial \Omega} \rho_w \left( \vec{u}_w \cdot \vec{dS} \right)= \rho_w^{ST}\left(Q_{we}^{ST} + Q_{wi}^{ST}-Q_w^{ST}\right) \, ;
\end{equation}

\item oil component:
\begin{equation}\label{eq: So2}
\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV = -\oint_{\partial \Omega} \rho_o Y_o \left(\vec{u}_o \cdot \vec{dS} \right) = \rho_o^{ST}Q_o^{ST} \, ;
\end{equation}

\item gas component:
\begin{equation}\label{eq: Sg2}
\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV = - \oint_{\partial \Omega} \left( \rho_o Y_g \vec{u}_o + \rho_g \vec{u}_g\right) \cdot \vec{dS} 
=  \rho_g^{ST}(Q_{gi}^{ST}-Q_g^{ST}) \, .
\end{equation}
\end{itemize}
%
%\begin{itemize}
%\item water component:
%\begin{equation}\label{eq: Sw2}
%\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV = \rho_w^{ST}Q_w^{ST} \, ;
%\end{equation}
%
%\item oil component:
%\begin{equation}\label{eq: So2}
%\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV = \rho_o^{ST}Q_o^{ST} \, ;
%\end{equation}
%
%\item gas component:
%\begin{equation}\label{eq: Sg2}
%\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV = \rho_g^{ST}Q_g^{ST} \, .
%\end{equation}
%\end{itemize}
%
\subsubsection{Averaging quantities}

@@

All quantities are initially functions of space and time and they become functions of time only after being averaged in space:

@@

\begin{itemize}
\item water component:
\begin{equation}\label{eq: Sw3}
\frac{\partial}{\partial t} \left( \bar{\rho_w}\bar{S}_w\bar{\phi} \Omega \right) = \rho_w^{ST}\left(Q_{we}^{ST} + Q_{wi}^{ST}-Q_w^{ST}\right)
\end{equation}

@@

\item gas component:
\begin{equation}\label{eq: Sg3}
\frac{\partial}{\partial t} \left[\left( \bar{\rho}_o\bar{Y}_g \bar{S}_o + \bar{\rho}_g \bar{S}_g \right) \bar{\phi} \Omega\right] =  \rho_g^{ST}(Q_{gi}^{ST}-Q_g^{ST}) \, .
\end{equation}
\end{itemize}

@@

\begin{equation}\label{eq: Sw4}
\frac{\partial}{\partial t} \left(\frac{\bar{S}_w V_p}{B_w} \right) = Q_{we}^{ST} + Q_{wi}^{ST}-Q_w^{ST}
\end{equation}

@@

\begin{equation}\label{eq: Sg4}
\frac{\partial}{\partial t} \left[ V_p \left(\frac{\bar{S}_g}{B_g} + \frac{\bar{S}_o R_s}{B_o} \right) \right] = \left( Q_{gi}^{ST}-Q_g^{ST} \right) \, ,
\end{equation}
